### 2021-07-05 rad-icfp-2021/generalization.tex
US spelling "Generalization"

\section{Generalisations}
\label{sec:generalisation}

@@

TODO: introduce sum and tensor of semantics of ground type\\
TODO: then * becomes a dot product\\
TODO: minor changes from there\\
TODO: except I'm not sure with reduce and shift1...

Though this has the right complexity, it is an interesting avenue for future research to find
better representations to allow for more optimizations. 
In particular, representations looking like Einsum \cite{van2011numpy} could be of interest 
and has been recently studied in the context of AD \cite{laue2018computing,laue2020simple}.

TODO: more generally about tensors \cite{liao2019differentiable,bernstein2020differentiating}

>>>>

\section{Generalizations}
\label{sec:generalization}

@@

A ground type $G$ is interpreted as an Euclidean space $A$. 
It is in particular a real vector space.
Similarly, a ground context $\Gamma=x1:G1,...,xn:Gn$ is interpreted as $\bigoplus_{1\leq i\leq n}Ai$, where $Ai\defeq\sem{Gi}$.
The denotation of the gradient of a term $\Gamma \vdash$!e: !$G$ at a point is then a matrix, more precisely an element of $()\bigoplus_{1\leq i\leq n}Ai)\otimes A$
where $\otimes$ is the tensor product of real vector spaces. This space is isomorphic to $\bigoplus_{1\leq i\leq n}Ai\otimes A$.

We can define $\otimes$ on the types of our language inductively by

\begin{tabular}{r c l}
    $\reals \otimes A$ & $\defeq$ & $A$ \\
    $(A1 \times ... \times An)\otimes A$ & $\defeq$ & $(A1\otimes A) \times ... \times (An \otimes A)$ \\
    $\Array{A1} \otimes A$ & $\defeq$ & $\Array{A1\otimes A}$
\end{tabular}

With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A1\times...\times An$ as expected.
TODO: define *, finish D(reduce)

We will need a new primitive $\widehat{*}$ whose denotation is this map.

Then, using this notation, there are only minimal changes to the reverse derivatives of !map2! and !reduce!.

\begin{center}
\begin{tabular}{r c l}
    $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !map2 (x,y.e1: G) e2 e3!) &=&  
    !let A,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
    && !let B,Y2 = !$\directD{\rho}{\Gamma,A}{Y1}$!(e3) in! \\
    && !let C=!$\grad_{\Gamma}$!e1 in!\\
    && !<map2 (x,y.e1) A B, fun (x1,...,xn,Z) -> !\\
    && !Y2( (x1,...,xn)!$\widehat{+}$!(C!$\widehat{*}$!(reduce + 0 Z)),!\\
    && \quad\quad! map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x]!$\widehat{*}$!b) A Z,!\\
    && \quad\quad! map2 (a,b.(!$\grad_{\{y\}}$!e1)[a/x]!$\widehat{*}$!b) B Z )>!\\
    \end{tabular}
    \end{center}

\begin{center}
\begin{tabular}{r c l}
        $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !reduce (x,y.e1) e2 e3!) 
        &=&  !let y1,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
        && !let A,Y2 = !$\directD{\rho}{\Gamma,y1}{Y1}$!(e3) in! \\
        && !let A0 = scanl (x,y.e1) y1 A in! \\
        && !let A1 = map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x,b/y]) A0 A! \\
        && !let A2 = shift1 (map2 (a,b.(!$\grad_{\{y\}}$!)e1[a/x,b/y]) A0 A)! \\
        && !let A3 = scanr * 1 A1! \\
        && !<reduce (x,y.e1) y1 A, fun (x1,...,xn,z) ->! \\
        && !Y((x1,...,xn),0,map2 (a,b. a*b*z) A2 A3)!
\end{tabular}
\end{center}

Evidently, one can combine all the generalizations from the previous subsections.
Even though this transformation has the correct complexity, it is open for future research to find
even better representations to allow for more optimizations. 
In particular, representations looking like Einsum \cite{van2011numpy} could be of interest 
and has been recently studied in the context of AD \cite{laue2018computing,laue2020simple}.
More generally, there is growing interest in tensor calculus \cite{liao2019differentiable,bernstein2020differentiating}.

### 2021-07-05 rad-icfp-2021/relwork.tex

We have introduced a transformation on programs to compute provably efficient (\ref{sec:complexity}) 
gradients via reverse derivatives in a purely functional way (\ref{sub:Macro for pure reverse mode transformation})
on a simple yet expressive language with functions on arrays (\ref{sub:sourcelang}), 
combined with standard functional optimizations (\ref{fig:optim}).  
We introduced a novel intermediate representation, unary form (\ref{sec:unf}) 
to decompose our transformation into simpler transformations.
We gave denotational semantics to our languages and our transformations (\ref{sec:correctness}) 
and showed correctness of the reverse-mode transformation (\ref{sub:Correctness theorem}).
We showed (\ref{sec:generalisation}) how to lift some of the restrictions that
we introduced on arrays and how to extend our approach to a richer language with more primitive operations and conditionals.

>>>>

We have introduced a transformation on programs to compute provably efficient (\ref{sec:complexity}) 
gradients via reverse derivatives in a purely functional way (\ref{sub:Macro for pure reverse mode transformation})
on a simple yet expressive language with functions on arrays (\ref{sub:sourcelang}), 
combined with standard functional optimizations (\ref{fig:optim}).  
We introduced a novel intermediate representation, unary form (\ref{sec:unf}) 
to decompose our transformation into simpler transformations.
We gave denotational semantics to our languages and our transformations (\ref{sec:correctness}) 
and showed correctness of the reverse-mode transformation (\ref{sub:Correctness theorem}).
We showed (\ref{sec:generalization}) how to lift some of the restrictions that
we introduced on arrays and how to extend our approach to a richer language with more primitive operations and conditionals.
